Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2023-01-03 response_to_reviewers_R1.Rmd

> This is an artifact of the attempt to track changes and does not appear in the unannotated manuscript.

- Section 6.2, line 52, Figure 6 should be Figure 7; likewise the figure number in the next sentence

> This is a downstream consequence of the previous error with Figure ??

- Figure 6, in my version the images are stacked rather than left/right

- Section 6.2, line 46 This sentence that starts “Consider, for example, “ seems out of place with the rest of the tone of this article

> TODO

- Section 7, line 7 or 8, it might be worth mentioning the growth of small conferences like UCSAS that have sprung up, something like “As the field of sports analytics has grown, there has been a proliferation of sports specific and regional sports analytics conferences.”

### 2023-01-03 sports_analytics.Rmd
add chess figures; #44

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And of course in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].

@@

Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
At the same time, many sports are similar, either because one evolved from the other, or the structure of the game shares certain attributes. 
Sports that are closely related anthropologically may or may not share common applications of analytical methods. 
For example, despite belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different.
For instance, an Elo rating could be equally valid for chess players and ice hockey teams.

@@

@kumagai2021hockey offer an EPV metric for ice hockey.

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package). 
We note how the contextual information provided by the chessboard is instrumental in helping the reader contextualize the data. 
(How many of us can visualize PGN directly?)
In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.

```{r chess, eval=whoami::username() == 'bbaumer', fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
game() |> plot()

@@

m <- extract_moves_as_game(greg[6, ])
# plot_moves(m, interactive = FALSE)
plot(m)
```

@@

The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
Many influential results have been showcased for the first time at these conferences. 
Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).

>>>>

Insights derived from the analysis of data have transformed the world of sports over the last few decades. 
While baseball---a naturally discrete sport with more than a century's worth of professional data---may be the sport with the longest relationship with sports analytics, one would be hard-pressed to identify a professional sport today in which sports analytics is not having an impact. 
In basketball, analytics has driven a shift in the conventional wisdom about shot selection. 
Most teams are shooting more three-pointers, settling for fewer long two-point shots, deploying more versatile defenders, and relying less on the strategy of pounding the ball into the paint in an attempt to get a high-percentage shot [@Schuhmann2021nba3point].
In American football, teams are going for it on fourth down far more often than in the past, a direct result of statistical analysis showing that most teams were previously overly conservative [@lopez2020bigger].
And, of course, in baseball, teams are using defensive shifts to maximize the probability of recording an out, encouraging hitters to improve their launch angles, and optimizing pitcher repertoires to minimize contact [@healey2017new].

@@

Every sport is different, with its own set of rules, strategies, methods of data collection, number of players, and the magnitude of the role of chance. 
At the same time, many sports are similar, either because one evolved from the other, or the structure of the games share certain attributes. 
Sports that are closely related historically may or may not share common applications of analytical methods. 
For example, despite belonging to the same bat-and-ball family, baseball and cricket differ in strategies such as batting order or sacrifice plays.
Conversely, with just a few small tweaks, analytical metrics might work just as well across sports that are unrelated and quite different.
For instance, an Elo rating could be equally valid for chess players and ice hockey teams.

@@

@macdonald2012expected uses expected goals to evaluate ice hockey players, but does not have access to player tracking data necessary to evaluate possessions.
@kumagai2021hockey offer an EPV metric for ice hockey.

@@

Figure \ref{fig:chess} shows a rendering of the starting chess board obtained via the \pkg{chess} package, along with the final position in the game won by one of the authors mentioned earlier in Section \ref{sec:elo} (with data downloaded via the \pkg{chessR} package). 
We note how the contextual information provided by the chessboard is instrumental in helping the reader understand the data. 
(How many of us can visualize PGN directly?)
In Section \ref{sec:graphics}, we outline a collection of graphical tools that provide similar context for different playing surfaces.

```{r chess-fig, eval=whoami::username() == 'bsbaumer'}
library(chess)
# Sys.setenv(RETICULATE_PYTHON = "my_env/bin/python3")
pdf("chess-1.pdf")
game() |> plot()
dev.off()

@@

m <- extract_moves_as_game(greg[6, ])
# plot_moves(m, interactive = FALSE)

pdf("chess-2.pdf")
plot(m)
dev.off()
```

```{r chess, fig.show="hold", out.width="49%", fig.cap="At left, the starting chess board printed via the \\pkg{chess} package. At right, the final position for one of the authors' recent wins (a checkmate playing Black)."}
knitr::include_graphics("chess-1.pdf")
knitr::include_graphics("chess-2.pdf")
```

@@

As the field of sports analytics has grown, a proliferation of sports specific and regional sports analytics conferences have arisen.
The biennial [New England Symposium on Statistics in Sports](https://www.nessis.org/) is likely the longest-running academic conference devoted to sports analytics. 
It's West Coast counterpart is [The Cascadia Symposium on Statistics in Sports](http://www.cascadiasports.com/). 
Many influential results have been showcased for the first time at these conferences. 
Other prominent sports analytics conferences include the  [Carnegie Mellon Sports Analytics Conference](https://www.stat.cmu.edu/cmsac/conference/2022/), [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2022/), and [MathSport International](http://www.mathsportinternational.com/).
